---
title: "5-Data transformation"
author: "Camilo Mayorquín"
date: "March 7, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# nycflights13

```{r }
library(nycflights13)
library(tidyverse)
flights
```

# The power of dplyr

Pick observations by their values - filter
Reorder rows - arrange
Pick variables by their names - select
Create new variables with functions of existing variables - mutate
Collapse many values down to a single summry - summarise
group_by

# Filter rows

```{r}
filter(flights, month == 1, day == 1)
# This returns a new data frame
```

# Comparisons

```{r}
sqrt(2) ^ 2 == 2
near(sqrt(2) ^ 2, 2)
# R is too precise sometimes, use near()
```

# Logical opertator
& 'and'
| 'or'
! 'not'

```{r}
filter(flights, month == 11 | month == 12)

# Using this evaluates to TRUE in a numeric context
filter(flights, month == 11 | 12)

# Instead, use this short-hand
nov_dec <- filter(flights, month %in% c(11,12))
# Selects every row where x (month) is one of the values en y (11,12)
```

```{r}
# Using De Morgan's law
# flights that weren't delayed on arrival or departure by more than two hours
filter(flights, dep_delay <=120 , arr_delay <= 120)
filter(flights, !(dep_delay > 120 | arr_delay > 120))
```

# Missing values

is.na() to determinate is a value is missing
filter() excludes both FALSE and NA

```{r}
df <- tibble(x = c(1, NA, 3))
filter(df, x > 1)
# If NAs are to be kept, you need to ask for them explicity
filter(df, is.na(x) | x > 1)
```

# Exercises

```{r}
# Flights had an arrival delay of 2 or more hours
filter(flights, arr_delay >= 120)
filter(flights, !(arr_delay < 120))
```

```{r}
# Flights to Houston (IAH or HOU)
filter(flights, dest %in% c('IAH','HOU'))
filter(flights, dest == 'IAH' | dest == 'HOU'  )
```

```{r}
# Operated by United, American or Delta
filter(flights, carrier %in% c('UA','AA','DL'))
```

```{r}
# Departed in summer (July, August, September)
filter(flights, month %in% c(7,8,9))
```

```{r}
# Arrived more than two hours late, but didn't leave late
filter(flights, arr_delay > 120 & dep_delay <=0)
```

```{r}
# Either both delay and more that 30 min flight time
filter(flights, (arr_delay >= 60 | dep_delay >= 60) & air_time > 30)
```

```{r}
# Departed between midnight and 6am (inclusive)
filter(flights, hour >=6 & hour <= 24)
filter(flights, between(x = hour,left = 6,right = 24))
```

```{r}
# Files with a missing dep_time
filter(flights, is.na(dep_time))
```

# Arrange rows

```{r}
# Can organize with multiple columns
arrange(flights, year, month, day)
```

```{r}
# Organize in descending order
arrange(flights,desc(dep_delay))
```

Missing values are always sorted at the end

# Exercises

```{r}
# All missing values at the start
arrange(flights, desc(is.na(dep_delay)))
```

```{r}
# most delayed flights
arrange(flights, desc(arr_delay),desc(dep_delay))
# least delayed flights
arrange(flights, arr_delay,dep_delay)
```

```{r}
# Fastest flights
arrange(flights, air_time)
```

# Select columns

Narrow to the variables wanted. Subset the data.

```{r}
# General idea
select(flights, year, month, day)
# All columns between year and date
select(flights, year:day)
# All columns except those from year to day (inclusive)
select(flights, -(year:day))
```

Helper functions that can be used with select() :

starts_with('abc')
ends_with('xyz')
constains('ijk')
matches('(.)\\1') regular expression for matching any variables that contain repeated characters
num_range('x', 1:3) : matches x1, x2 and x3

select() can rename variables but rename() keeps all the variables not explicitly mencioned

```{r}
rename(flights, tail_num = tailnum)
```

everything() is useful for moving certain variables to the start of the data frame

```{r}
select(flights, time_hour, air_time, everything())
```

# EXERCISES

```{r}

# one_of evaluates() if the vector names correspond to actual variable names and selects them, if not, throws a warning but no error
vars <- c("year", "month", "day", "dep_delay", "arr_delay")

select(flights, one_of(vars))

```

```{r}
# Select anything that contains the string in its name, no matter if caps are present
select(flights, contains("TIME"))
```

