---
title: "7-Exploratory Data Analysis"
author: "Camilo Mayorquín"
date: "6/17/2019"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
```


# Introduction

- What type of variation occurs within my variables?
- What type of covariation occurs between my variables?

# Variation

Variation is the tendency of the values of a variable to change from measurement from measurement.

# Visualising distributions

To examine the distribution of a categorical variable use a bar chart
```{r}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut))
```

You can compute the values manually with dplyr

```{r}
diamonds %>% 
  count(cut)
```

To examine the distribution of a continuous variable use a histogram

```{r}
ggplot(data = diamonds) +
  geom_histogram(mapping = aes(x = carat), binwidth = 0.5)
```

Can also be computed manually

```{r}
diamonds %>% 
  count(cut_width(carat, 0.5))
```

Zoom in to see different pattern

```{r}
smaller <- diamonds %>% 
  filter(carat < 3)

ggplot(data = smaller, mapping = aes(x = carat)) + geom_histogram(binwidth = 0.1)
```

To overlay multiple histograms is better to use geom_freqpoly as it does the same calculation

```{r}
ggplot(data = diamonds, mapping = aes(x = carat, colour = cut)) +
  geom_freqpoly(binwidth = 0.1)
```

Search for patterns

This plot for example appears to contain clusters
```{r}
ggplot(data = smaller, mapping = aes(x = carat)) +
  geom_histogram(binwidth = 0.01)
```

Why two clusters in this plot?
```{r}
ggplot(data = faithful, mapping = aes(x = eruptions)) + 
  geom_histogram(binwidth = 0.25)
```

Gotta check relationship between variables

# Unusual values

This plot shows an unusual wide limit for x axis
```{r}
ggplot(diamonds) +
  geom_histogram(mapping = aes(x = y), binwidth = 0.5)
```

Zoom in to small values of the y axis to see

```{r}
ggplot(diamonds) + 
  geom_histogram(mapping = aes(x = y), binwidth = 0.5) +
  coord_cartesian(ylim = c(0,50))
```

We see there are unusual values of 0, ~30, ~60, we pluck them out with dplyr

```{r}
unusual <- diamonds %>% 
  filter(y < 3 | y > 20) %>% 
  select(price, x, y, z) %>% 
  arrange(y)
unusual
# Diamonds with 0mm dimension, it must be incorrect
```

# Excercises

Explore price distribution
```{r}
ggplot(diamonds) + 
  geom_histogram(mapping = aes(x = price), binwidth = 100) +
  coord_cartesian(ylim = c(0,200))
```

Count diamonds with 0.99 carats
```{r}
diamonds %>%
  filter(carat == 0.99) %>% 
  count()
```

# Missing values

Dropping the entire row with strange values is not recommended. Instead replace these with missing values

```{r}
diamonds2 <- diamonds %>% 
  # If y between this values replace with NA
  # Else keep the same value
  mutate(y = ifelse(y<3 | y>20, NA, y))
```

ggplot does not include missing values in the plot, but it warns that they have been removed

```{r}
ggplot(data = diamonds2, mapping = aes(x = x, y = y)) +
  geom_point(na.rm=T)
```

In the flights data set, missing values for the dep_time variable indicate that the flight was cancelled. Let's compare the scheduled departed times for cancelled and non-cancelled times.

```{r}
flights %>% 
  mutate(
    cancelled = is.na(dep_time),
    sched_hour = sched_dep_time %/% 100,
    sched_min = sched_dep_time %% 100,
    sched_dep_time = sched_hour + sched_min / 60
  ) %>% 
  ggplot(mapping = aes(sched_dep_time)) +
  geom_freqpoly(mapping = aes(colour = cancelled), binwidth = 1/4)
```

Excercises


Missing values do not appear in barplot since they have no frequency
```{r}
flights %>% 
  mutate(
    cancelled = is.na(dep_time),
    sched_hour = sched_dep_time %/% 100,
    sched_min = sched_dep_time %% 100,
    sched_dep_time = sched_hour + sched_min / 60
  ) %>% 
  ggplot(mapping = aes(sched_dep_time)) +
  geom_bar(mapping = aes(colour = cancelled))
```


# Covariation

Describes the variation between variables. Is the tendency for the values of two or more variables to vary together in a related way.

Categorical and continuous variable

See the comparison between this type of variables using a frequency polygon makes it hard to see the difference if some groups are much smaller than others. For example:

```{r}
ggplot(data = diamonds, mapping = aes(x = price)) +
  geom_freqpoly(mapping = aes(colour = cut), binwidth = 500)
```

It's hard to see the difference in distribution because the overall counts differ so much:

```{r}
ggplot(diamonds) + 
  geom_bar(mapping = aes(x = cut))
```

To make the comparison easier we need to swap what is displayed on the y-axis. Instead of displaying count, we’ll display density, which is the count standardised so that the area under each frequency polygon is one.

```{r}
ggplot(data = diamonds, mapping = aes(x = price, y = ..density..)) +
  geom_freqpoly(mapping = aes(colour = cut), binwidth = 500)
```

it appears that fair diamonds (the lowest quality) have the highest average price! But maybe that’s because frequency polygons are a little hard to interpret - there’s a lot going on in this plot.

Another alternative is the BOXPLOT: gives a sense of the spread of the distribution and whether or not the distribution is symmetric about the median or skewed to one side.

Distribution of price by cut using geom_boxplot():

```{r}
ggplot(data = diamonds, mapping = aes(x = cut, y = price)) +
  geom_boxplot()
```

We can see that better quality diamonds are cheaper on average!

cut is an ordered factor. If you have a variable without such an intrinsic order, you might want to reorder them to make a more informative display.

Let's see that with the class variable in the "mpg" dataset

```{r}
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) +
  geom_boxplot()
```

To make the trend easier to read, we can reorder class based on the median value of hwy:

```{r}
ggplot(data = mpg) +
  geom_boxplot(mapping = aes(x = reorder(class, hwy, FUN=median), y = hwy)) +
  # For long name variables
  coord_flip()
```

Excercises

Use geom_boxploth() function from ggstance package and compare it to coord_flip()

It produces the same plot and the axis are more intuitive
```{r}
library(ggstance)
ggplot(data = mpg) +
  geom_boxploth(aes(x = hwy,y = reorder(class, hwy, FUN=median)))
```

One problem with boxplots is that they were developed in an era of much smaller datasets and tend to display a prohibitively large number of “outlying values”.

Install the lvplot package and use the geom_lv() function to display the distribution of price vs cut

```{r}
library(lvplot)
ggplot(data = diamonds, aes(x = cut, y = price)) +
  geom_lv()
```

Today I learned something important: For datasets n < 1000 the use of boxplots is fine, but for large size data sets, 10000 < n < 100000 it's more convenient to use a letter value box plot. 

```{r}
ggplot(data = diamonds, aes(x = cut, y = price)) +
  geom_violin()
```

For samller datasets, the relationship between a continous and categorical variable it's sometimes better seen with geom_jitter() or the functions provided by the ggbeeswarm package 

```{r}
library(ggbeeswarm)
ggplot(data = diamonds, aes(x = cut, y = price)) +
  geom_beeswarm(aes(color=cut))
```

