---
title: "7-Exploratory Data Analysis"
author: "Camilo Mayorquín"
date: "6/17/2019"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
```


# Introduction

- What type of variation occurs within my variables?
- What type of covariation occurs between my variables?

# Variation

Variation is the tendency of the values of a variable to change from measurement from measurement.

# Visualising distributions

To examine the distribution of a categorical variable use a bar chart
```{r}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut))
```

You can compute the values manually with dplyr

```{r}
diamonds %>% 
  count(cut)
```

To examine the distribution of a continuous variable use a histogram

```{r}
ggplot(data = diamonds) +
  geom_histogram(mapping = aes(x = carat), binwidth = 0.5)
```

Can also be computed manually

```{r}
diamonds %>% 
  count(cut_width(carat, 0.5))
```

Zoom in to see different pattern

```{r}
smaller <- diamonds %>% 
  filter(carat < 3)

ggplot(data = smaller, mapping = aes(x = carat)) + geom_histogram(binwidth = 0.1)
```

To overlay multiple histograms is better to use geom_freqpoly as it does the same calculation

```{r}
ggplot(data = diamonds, mapping = aes(x = carat, colour = cut)) +
  geom_freqpoly(binwidth = 0.1)
```

Search for patterns

This plot for example appears to contain clusters
```{r}
ggplot(data = smaller, mapping = aes(x = carat)) +
  geom_histogram(binwidth = 0.01)
```

Why two clusters in this plot?
```{r}
ggplot(data = faithful, mapping = aes(x = eruptions)) + 
  geom_histogram(binwidth = 0.25)
```

Gotta check relationship between variables

# Unusual values

This plot shows an unusual wide limit for x axis
```{r}
ggplot(diamonds) +
  geom_histogram(mapping = aes(x = y), binwidth = 0.5)
```

Zoom in to small values of the y axis to see

```{r}
ggplot(diamonds) + 
  geom_histogram(mapping = aes(x = y), binwidth = 0.5) +
  coord_cartesian(ylim = c(0,50))
```

We see there are unusual values of 0, ~30, ~60, we pluck them out with dplyr

```{r}
unusual <- diamonds %>% 
  filter(y < 3 | y > 20) %>% 
  select(price, x, y, z) %>% 
  arrange(y)
unusual
# Diamonds with 0mm dimension, it must be incorrect
```

# Excercises

Explore price distribution
```{r}
ggplot(diamonds) + 
  geom_histogram(mapping = aes(x = price), binwidth = 100) +
  coord_cartesian(ylim = c(0,200))
```

Count diamonds with 0.99 carats
```{r}
diamonds %>%
  filter(carat == 0.99) %>% 
  count()
```

# Missing values
